---
title: "gdm_model_out"
author: "Rob Cooke"
date: "08/06/2021"
output: html_notebook
---

## Load model ##

```{r}

##############
# chose taxonomic group
tax <- "birds"
##############

# load model
gdmr <- readRDS(paste0("data/", tax, "/out_data/", tax, "_gdmr.rds"))

# reference cells as comparison - 20% of cells
ref_cells <- readRDS("data/ref_cells.rds")

# transformed predictors current
# tmin is dropped for butterflies as the coefficient = 0
tr_curr <- readRDS(paste0("data/", tax, "/out_data/", tax, "_tr_curr.rds"))

# transformed predictors current - reference cells - 20% of cells
tr_curr_rc <- readRDS(paste0("data/", tax, "/out_data/", tax, "_tr_curr_rc.rds"))

## helper functions

# function to remove intercept from gdm predictions - see Mokany et al., 2022
remove_int <- function(mod, pred){
  adjust <- 0 - log(1 - pred) - mod$intercept
  
  adjust_diss <- 1 - exp(0 - adjust)
  
  return(adjust_diss)
}

# function to exit function
exit <- function() { invokeRestart("abort") }  

# first letter to uppercase
firstup <- function(x) {
  substr(x, 1, 1) <- toupper(substr(x, 1, 1))
  x
}

# assumed species area relationship following Di Marco et al (2019) and references contained within
z <- 0.25

# baseline year
baseline_yr <- 2015

```

## Disappearing and novel environments ##

adapted code from Mokany et al., 2022

```{r}

#### disappearing ####

# current covariates
covs_dis <- as.data.frame(raster::stack(covars), xy = TRUE) %>% 
  # drop nas
  tidyr::drop_na() %>%
  # rename x and y for predict.gdm function
  dplyr::rename(xCoord = x, yCoord = y) %>% 
  # add s1 to column names
  dplyr::rename_with(.cols = dplyr::everything(), function(x) {paste0("s1.", x)}) %>% 
  # distance and weights for predict.gdm function
  dplyr::mutate(distance = 1,
                weights = 1)

# function to calculate disappearing environments per scenario
disap <- function(covs_dis, gdmr, scenario, scenario_name, tax) {
  
  # future covariates
  fut_covs <- as.data.frame(raster::stack(scenario), xy = TRUE) %>% 
    # drop nas
    tidyr::drop_na() %>% 
    # rename x and y for predict.gdm function
    dplyr::rename(xCoord = x, yCoord = y) %>% 
    # add s2 to column names
    dplyr::rename_with(.cols = dplyr::everything(), function(x) {paste0("s2.", x)})
  
  print("step 1: done")
  
  # calculation of disappearing environments
  disappear_gdm <- pbmcapply::pbmclapply(1:nrow(covs_dis), function(i) {
    
    # focal cell
    one_cell <- covs_dis[i,]
    
    # join focal cell to all future cells
    dat <- dplyr::bind_cols(one_cell, fut_covs) %>% 
      # reorder columns for predict.gdm function
      dplyr::select(distance, weights, s1.xCoord, s1.yCoord, s2.xCoord, s2.yCoord, dplyr::starts_with("s1"), dplyr::starts_with("s2"))
    
    # predict dissimilarity between current cell and all future cells
    pred_diss <- predict(object = gdmr, dat, time = FALSE)
    
    # remove effect of intercept
    # see Mokany et al., 2022
    pred_diss_adj <- remove_int(gdmr, pred_diss)
    
    # combine into dataframe (min, mean, median)
    pred_diss_df <- data.frame(min = min(pred_diss_adj), mean = mean(pred_diss_adj), median = median(pred_diss_adj))
    
  }, mc.cores = 4, mc.preschedule = FALSE) %>% 
    dplyr::bind_rows()
  
  print("step 2: done")
  
  disappear_df <- dplyr::bind_cols(dplyr::select(covs_dis, s1.xCoord, s1.yCoord), disappear_gdm) %>% 
    setNames(c("x", "y", "disappearing_min", "disappearing_mean", "disappearing_median"))
  
  print("step 3: done")
  
  saveRDS(disappear_df, paste0("data/", tax, "/out_data/disappear/", tax, "_disappear_", scenario_name, ".rds"))

  print("finished")
  
}

disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp26_2030, scenario_name = "rcp26_2030", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp26_2050, scenario_name = "rcp26_2050", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp26_2070, scenario_name = "rcp26_2070", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp45_2030, scenario_name = "rcp45_2030", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp45_2050, scenario_name = "rcp45_2050", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp45_2070, scenario_name = "rcp45_2070", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp60_2030, scenario_name = "rcp60_2030", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp60_2050, scenario_name = "rcp60_2050", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp60_2070, scenario_name = "rcp60_2070", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp85_2030, scenario_name = "rcp85_2030", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp85_2050, scenario_name = "rcp85_2050", tax = tax)
disap(covs_dis = covs_dis, gdmr = gdmr, scenario = covars_rcp85_2070, scenario_name = "rcp85_2070", tax = tax)

#### novel ####

# takes ~10 hours per scenario

# current covariates
covs_nov <- as.data.frame(raster::stack(covars), xy = TRUE) %>% 
  # drop nas
  tidyr::drop_na() %>%
  # rename x and y for predict.gdm function
  dplyr::rename(xCoord = x, yCoord = y) %>% 
  # add s2 to column names
  dplyr::rename_with(.cols = dplyr::everything(), function(x) {paste0("s2.", x)})

# function to calculate novel environments per scenario
novel <- function(covs_nov, gdmr, scenario, scenario_name, tax) {
  
  # future covariates
  fut_covs <- as.data.frame(raster::stack(scenario), xy = TRUE) %>% 
    # drop nas
    tidyr::drop_na() %>% 
    # rename x and y for predict.gdm function
    dplyr::rename(xCoord = x, yCoord = y) %>% 
    # add s2 to column names
    dplyr::rename_with(.cols = dplyr::everything(), function(x) {paste0("s1.", x)}) %>%
    # distance and weights for predict.gdm function
    dplyr::mutate(distance = 1,
                  weights = 1)
  
  print("step 1: done")
  
  # calculation of novel environments
  novel_gdm <- pbmcapply::pbmclapply(1:nrow(fut_covs), function(i) {
    
    # focal future cell
    one_cell <- fut_covs[i,]
    
    # join focal future cell to all current cells
    dat <- dplyr::bind_cols(one_cell, covs_nov) %>% 
      # reorder columns for predict.gdm function
      dplyr::select(distance, weights, s1.xCoord, s1.yCoord, s2.xCoord, s2.yCoord, dplyr::starts_with("s1"), dplyr::starts_with("s2"))
    
    # predict dissimilarity between future cell and all current cells
    pred_diss <- predict(object = gdmr, dat, time = FALSE)
    
    # remove effect of intercept
    # see Mokany et al., 2022
    pred_diss_adj <- remove_int(gdmr, pred_diss)
    
    # combine into dataframe (min, mean, median)
    pred_diss_df <- data.frame(min = min(pred_diss_adj), mean = mean(pred_diss_adj), median = median(pred_diss_adj))
    
  }, mc.cores = 4, mc.preschedule = FALSE) %>% 
    dplyr::bind_rows()
  
  print("step 2: done")
  
  novel_df <- dplyr::bind_cols(dplyr::select(fut_covs, s1.xCoord, s1.yCoord), novel_gdm) %>% 
    setNames(c("x", "y", "novel_min", "novel_mean", "novel_median"))
  
  print("step 3: done")
  
  saveRDS(novel_df, paste0("data/", tax, "/out_data/novel/", tax, "_novel_", scenario_name, ".rds"))

  print("finished")
  
}

novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp26_2030, scenario_name = "rcp26_2030", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp26_2050, scenario_name = "rcp26_2050", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp26_2070, scenario_name = "rcp26_2070", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp45_2030, scenario_name = "rcp45_2030", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp45_2050, scenario_name = "rcp45_2050", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp45_2070, scenario_name = "rcp45_2070", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp60_2030, scenario_name = "rcp60_2030", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp60_2050, scenario_name = "rcp60_2050", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp60_2070, scenario_name = "rcp60_2070", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp85_2030, scenario_name = "rcp85_2030", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp85_2050, scenario_name = "rcp85_2050", tax = tax)
novel(covs_nov = covs_nov, gdmr = gdmr, scenario = covars_rcp85_2070, scenario_name = "rcp85_2070", tax = tax)

```

## Species committed to extinction under land-use change and climate change ##

adapted from code in Mokany et al., 2022 and equations in Di Marco et al., 2019

use this function to create tr_hc_fut_rc for each future

```{r}

# function to calculate persistence under land-use change and climate change
pers <- function(gdmr, tr_curr, tr_curr_rc, ref_cells, hc_scenario, cc_scenario, scenario_name, tax, save = FALSE) {

  # extract habitat condition data from raster with x and y coordinates
  hc_df <- as.data.frame(hc_scenario, xy = TRUE) %>%
    # rename habitat condition score as "hc"
    dplyr::rename(hc = mean)

  # transform future environmental covariates based on ecological relationships
  # a non-linear transform of each environmental grid to include the estimated coefficients
  # see Manion G., (2009) A technique for monotonic regression splines to enable non-linear transformation of GIS rasters. 18th World IMACS/MODSIM Congress, Cairns, Australia. https://www.mssanz.org.au/modsim09/F13/manion_F13a.pdf
  trans_fut <- gdm::gdm.transform(model = gdmr, data = raster::stack(cc_scenario)) %>%
    # convert to dataframe with x and y coordinates
    as.data.frame(xy = TRUE)

  # join transformed future environmental covariates and habitat condition data
  tr_hc_fut_rc <- dplyr::inner_join(trans_fut, hc_df, by = c("x", "y")) %>%
    # drop nas
    tidyr::drop_na() %>%
    # only reference cells - random 20% of cells
    dplyr::right_join(ref_cells, by = c("x", "y"))

  # save files for HPC
  if(save == TRUE) {

    # save tr_hc_fut_rc
    saveRDS(tr_hc_fut_rc, paste0("data/", tax, "/out_data/tr_hc_fut/", tax, "_", scenario_name, "_tr_hc_fut_rc.rds"))

    # exit function
    exit()

    }

  # calculate persistence values for each focal cell i for all reference cells j
  # run in parallel across 4 cores
  pers_df <- lapply(1:1, function(i) {
    # pbmcapply::pbmclapply

    # focal cell
    one_cell <- tr_curr[i,] # current

    # reference cells j
    rc_curr <- tr_curr_rc[-i,] # current
    rc_fut <- tr_hc_fut_rc[-i,] # future

    # similarity between focal cell i and current reference cells j(1:(0.2*n))
    similarity_ij_curr <- sapply(1:nrow(rc_curr), function(j) {

      # ecological distance between focal cell i and current reference cell j
      ecol_dist_curr <- sum(abs(as.matrix(dplyr::select(one_cell, xCoord:sun)) - as.matrix(dplyr::select(rc_curr[j,], xCoord:sun))))

      # compositional similarity between focal cell i and current reference cell j
      similarity_curr <- (exp(-1 * (gdmr$intercept + ecol_dist_curr)))

    }) # close sapply

    # summed similarity across current reference cells assume pristine habitat everywhere
    # Aibaseline in Di Marco et al., 2019 equation 1
    # denominator in Di Marco et al., 2019 equation 4
    sum_similarity_pristine <- sum(similarity_ij_curr)

    # similarity between focal cell i and future reference cells j(1:(0.2*n))
    similarity_ij_fut <- sapply(1:nrow(rc_fut), function(j) {

      # ecological distance between focal cell i and future reference cell j
      ecol_dist_fut <- sum(abs(as.matrix(dplyr::select(one_cell, xCoord:sun)) - as.matrix(dplyr::select(rc_fut[j,], xCoord:sun))))

      # compositional similarity between focal cell i and future reference cell j
      similarity_fut <- (exp(-1 * (gdmr$intercept + ecol_dist_fut)))

    }) # close sapply

    # summed similarity across future reference cells when accounting for habitat condition
    # numerator in Di Marco et al., 2019 equation 4
    sum_similarity_hc <- sum(similarity_ij_fut * as.matrix(dplyr::select(rc_fut, hc))[,1])

    # compile outputs and calculate persistence per cell
    pers_ij <- one_cell %>%
      dplyr::mutate(sim_hc = sum_similarity_hc,
                    sim_pris = sum_similarity_pristine) %>%
      # expected persistence using species-area relationship
      # pitest in Di Marco et al., 2019 equation 3
      dplyr::mutate(pers = (sim_hc / sim_pris) ^ z) %>%
      # extinction risk
      dplyr::mutate(extinct = 1 - pers) %>%
      # cell weight
      # w in Di Marco et al., 2019 equation 6
      dplyr::mutate(cell_weight = 1 / sim_pris) %>%
      # numermouldenator see Mokany et al., 2022 Supporting Information Appendix S1 - section 2.10
      dplyr::mutate(numerator = pers * cell_weight) %>%
      # numerator in geometric mean see Di Marco et al., 2019 equation 5
      dplyr::mutate(numerator_ln = log(pers) * cell_weight) %>%
      # select necessary columns
      dplyr::select(x, y, hc:numerator_ln)

  }) %>%
  #, mc.cores = 4, mc.preschedule = TRUE) %>%
    # bind across focal cells i
    dplyr::bind_rows()

  # # save dataframe
  # saveRDS(pers_df, paste0("data/", tax, "/out_data/", tax, "_pers_lu_", scenario_name, ".rds"))

}

# run function
# change future in three places
pers(gdmr = gdmr, tr_curr = tr_curr, tr_curr_rc = tr_curr_rc, ref_cells = ref_cells, hc_scenario = hc_rcp85_ssp5_2070, cc_scenario = covars_rcp85_2070, scenario_name = "rcp85_ssp5_2070", tax = tax, save = TRUE)

```

## Baseline ##

use this to create baseline_2020_tr_hc_rd

```{r}

# function to calculate persistence under land-use change only
baseline <- function(gdmr, tr_curr, ref_cells, hc_scenario, scenario_name, tax) {
  
  # extract habitat condition data from raster with x and y coordinates
  hc_df <- as.data.frame(hc_scenario, xy = TRUE) %>% 
    # rename habitat condition score as "hc"
    dplyr::rename(hc = LandUseIndex)
  
  # join transformed environmental covariates and habitat condition data
  tr_hc_rc <- dplyr::inner_join(dplyr::select(tr_curr, -hc), hc_df, by = c("x", "y")) %>% 
    # drop nas
    tidyr::drop_na() %>%
    # only refence cells
    dplyr::right_join(ref_cells, by = c("x", "y"))
  
  # save tr_hc
  saveRDS(tr_hc_rc, paste0("data/", tax, "/out_data/", tax, "_", scenario_name, "_tr_hc_rc.rds"))
  
}

# # run baseline function
# baseline(gdmr = gdmr, tr_curr = tr_curr, ref_cells = ref_cells, hc_scenario = hc_baseline_2020, scenario_name = "baseline_2020", tax = tax)

```

## Regional estimates of species committed to extinction under land-use change and climate change ##

```{r}

# function to calculate regional estimates of species committed to extinction
pers_reg <- function(scenario_name, tax, lims, leg = FALSE) {
  
  # load data
  pers_df <- readRDS(paste0("data/", tax, "/out_data/pers/", tax, "_", scenario_name, ".rds")) %>%   
    # expected persistence using species-area relationship
    # pitest in Di Marco et al., 2019 equation 3
    dplyr::mutate(pers = (sim_hc / sim_pris) ^ z) %>% 
    # extinction risk
    dplyr::mutate(extinct = 1 - pers) %>% 
    # cell weight
    # w in Di Marco et al., 2019 equation 6
    dplyr::mutate(cell_weight = 1 / sim_pris) %>% 
    # numerator see Mokany et al., 2022 Supporting Information Appendix S1 - section 2.10
    dplyr::mutate(numerator = pers * cell_weight) %>% 
    # numerator in geometric mean see Di Marco et al., 2019 equation 5
    dplyr::mutate(numerator_ln = log(pers) * cell_weight) %>% 
    # select necessary columns
    dplyr::select(x, y, hc:numerator_ln)
    
  # regional estimate of species committed to extinction
  ext_regional <- 1 - exp(sum(pers_df$numerator_ln) / sum(pers_df$cell_weight))
  
  splt <- unlist(strsplit(scenario_name, "_"))
  
  # dataframe for regional estimate
  if(splt[[1]] != "baseline") {
    ext_reg_df <- data.frame(tax = tax, rcp = as.character(gsub("\\D", "", splt[[1]])), ssp = as.character(gsub("\\D", "", splt[[2]])), year = as.numeric(splt[[3]]), ext_reg = as.numeric(ext_regional))
  } else {
      ext_reg_df <- data.frame(tax = tax, rcp = splt[[1]], ssp = splt[[1]], year = baseline_yr, ext_reg = as.numeric(ext_regional))
      }
  
  # save regional estimate
  saveRDS(ext_reg_df, paste0("data/", tax, "/out_data/ext_reg/", tax, "_", scenario_name, ".rds"))
  
}

# scenario names
sc_nms <- c("rcp26_ssp1_2030", "rcp26_ssp1_2050", "rcp26_ssp1_2070", "rcp45_ssp2_2030", "rcp45_ssp2_2050", "rcp45_ssp2_2070", "rcp45_ssp4_2030", "rcp45_ssp4_2050", "rcp45_ssp4_2070", "rcp60_ssp3_2030", "rcp60_ssp3_2050", "rcp60_ssp3_2070", "rcp85_ssp2_2030", "rcp85_ssp2_2050", "rcp85_ssp2_2070", "rcp85_ssp5_2030", "rcp85_ssp5_2050", "rcp85_ssp5_2070")

lapply(sc_nms, function(scenario_name) pers_reg(scenario_name = scenario_name, tax = tax))

# baseline
pers_reg(scenario_name = "baseline_2020", tax = tax)

```

## Reference cells and tr_curr ##

```{r}

# # reference cells
#
# # extract habitat condition data from raster with x and y coordinates
# hc_df <- as.data.frame(hc_rcp85_ssp5_2070, xy = TRUE) %>%
#   # rename habitat condition score as "hc"
#   dplyr::rename(hc = mean)
#
# # join transformed environmental covariates and habitat condition data
# tr_hc <- dplyr::inner_join(trans_curr, hc_df, by = c("x", "y")) %>%
#   # drop nas
#   tidyr::drop_na()
#
# ref_cells <- dplyr::slice_sample(tr_hc, prop = 0.20, replace = FALSE) %>%
#   dplyr::select(x, y)
#
# saveRDS(ref_cells, "data/ref_cells.rds")

# # tr_curr, tr_curr_rc, tr_curr_split
# 
# # transform environmental covariates based on ecological relationships
# # a non-linear transform of each environmental grid to include the estimated coefficients
# # see Manion G., (2009) A technique for monotonic regression splines to enable non-linear transformation of GIS rasters. 18th World IMACS/MODSIM Congress, Cairns, Australia. https://www.mssanz.org.au/modsim09/F13/manion_F13a.pdf
# trans_curr <- gdm::gdm.transform(model = gdmr, data = raster::stack(covars)) %>%
#   # convert to dataframe with x and y coordinates
#   as.data.frame(xy = TRUE)
# 
# hc_df <- as.data.frame(hc_rcp26_ssp1_2030, xy = TRUE) %>% # all habitat condition rasters have same number of non-na cells
#   # rename habitat condition score as "hc"
#   dplyr::rename(hc = mean)
# 
# # join transformed environmental covariates and habitat condition data
# tr_curr <- dplyr::inner_join(trans_curr, hc_df, by = c("x", "y")) %>%
#   # drop nas
#   tidyr::drop_na()
# 
# saveRDS(tr_curr, paste0("data/", tax, "/out_data/", tax, "_tr_curr.rds"))
# 
# tr_curr_rc <- tr_curr %>%
#   # only reference cells
#   dplyr::right_join(ref_cells, by = c("x", "y"))
# 
# saveRDS(tr_curr_rc, paste0("data/", tax, "/out_data/", tax, "_tr_curr_rc.rds"))
# 
# tr_curr_split <- dplyr::mutate(tr_curr, grp = as.numeric(ggplot2::cut_number(1:nrow(tr_curr), 10000))) %>%
#   split(., f = .$grp)
# 
# saveRDS(tr_curr_split, paste0("data/", tax, "/out_data/", tax, "_tr_curr_split.rds"))

```


